Online Learning Framework

### 2014-01-05 lp/framework.tex
writeup with lp, sum of pointed sets

\title{Online Learning Framework}
\author{Jacob Steinhardt}
\begin{document}
\maketitle
We are interested in the following learning problem: our experts 
consist of vectors $w \in \sC_P$, where $\sC_P$ is some convex 
set; similarly, our adversary can play any move $z_t$ with 
$z_t \in \sC_D$. A typical example is $\sC_P$ is the $n$-dimensional 
simplex and $\sC_D$ is the unit $l^{\infty}$ ball. Our end goal is 
to obtain a regret bound of the following type:
\[ \sum_{t=1}^T w_t^Tz_t \leq \frac{\rho}{\epsilon} + \epsilon \sum_{t=1}^T |w|^T|z_t| + \sum_{t=1}^T w^Tz_t. \]
If we let $\sigma = \sup_{w \in \sC_P, z \in \sC_D} |w|^T|z|$, then the regret is bounded by
$\frac{\rho}{\epsilon} + \epsilon \sigma T$, which is equal to $2\sqrt{\rho \sigma T}$ for 
$\epsilon = \sqrt{\frac{\rho}{\sigma T}}$. This is the type of regret bound that is typical 
in most online learning settings. However, our bound can lead to stronger results in some 
situations.

We propose the following general style of algorithm for obtaining such 
bounds. The key innovation is an auxiliary variable $x_t$ that lets us 
relate $w_t^Tz_t$ and $w^Tz_t$. The updates are as follows:
\begin{align*}
x_{t+1} &= f(x_t, \eta z_t) \\
w_{t+1} &= \argmax_{w \in \sC_P} w^Tx_{t+1} - \psi(w) \\
\Phi_{t+1} &= w_{t+1}^Tx_{t+1} - \psi(w_{t+1})  \\ &= \psi^*(x_{t+1}).
\end{align*}

@@

\paragraph{Formal statement.} To turn these approximations into a rigorous bound, we will 
need $f$ and $\psi$ to satisfy the following properties:
\begin{itemize}
\item Convexity of $\psi^* \circ f$: $\psi^*(f(x,\eta z)) \leq \psi^*(x) - \eta w_x^Tz$, where $w_x = \partial \psi^*(x)$.
\item Smoothness of $f$: $w^T(x-\eta z - f(x,\eta z)) \leq \eta^2 \alpha |w|^T|z|$ for all $w$.
\end{itemize}
(For some intuition, any $f$ satisfying the two properties above 
must satisfy $f(x,0) = x$ and $f(x,\eta z) = x-\eta z + O(\eta^2)$.)
\begin{theorem}
\label{thm:main}
Suppose that $\psi^*$ and $f$ satisfy the two properties above. Then, 
for $\eta = \frac{\epsilon}{\alpha}$, we have the bound
\[ \sum_{t=1}^T w_t^Tz_t \leq \alpha \frac{\Phi_1 + \psi(w) - w^Tx_1}{\epsilon} + \epsilon \sum_{t=1}^T |w|^T|z_t| + \sum_{t=1}^T w^Tz_t. \]
\end{theorem}
\begin{proof}
By the first condition, we have
\begin{align*}
\Phi^{t+1} &= \psi^*(x_{t+1}) \\
 &= \psi^*(f(x_t, \eta z_t)) \\
 &\leq \psi^*(x_t) - \eta \left(\frac{\partial \psi^*}{\partial x_t}\right)^T z_t \\
 &= \Phi_t - \eta w_t^T z_t,
\end{align*}
which by induction implies that
\[ \Phi_{T+1} \leq \Phi_1 - \eta \sum_{t=1}^T w_t^Tz_t. \]
By the second condition, we have
\begin{align*}
w^Tx_{t+1} &= w^Tf(x, \eta z) \\
 &\geq w^Tx_t - \eta w^Tz_t - \eta^2 \alpha |w|^T|z_t|,
\end{align*}
which by induction implies that
\[ w^Tx_{T+1} \geq w^Tx_1 - \eta^2\alpha \sum_{t=1}^T |w|^T|z_t| - \eta \sum_{t=1}^T w^Tz_t. \]
Combining this with the first inequality yields
\begin{align*}
\Phi_1 - \eta \sum_{t=1}^T w_t^Tz_t &\geq \Phi_{T+1} \\
 &\geq w^Tx_{T+1} - \psi(w) \\
 &\geq w^Tx_1 - \psi(w) - \eta^2\alpha \sum_{t=1}^T |w|^T|z_t| - \eta \sum_{t=1}^T w^Tz_t.
\end{align*}
Re-arranging yields
\[ \sum_{t=1}^T w_t^Tz_t \leq \frac{\Phi_1 + \psi(w) - w^Tx_1}{\eta} + \eta\alpha \sum_{t=1}^T |w|^T|z_t| + \sum_{t=1}^T w^Tz_t, \]
which gives the desired result.
\end{proof}
\section{Examples}
\paragraph{Multiplicative updates.} If we let $\sC_P$ be the simplex and $\sC_D$ be the 
$l^{\infty}$ ball of radius $r$, then a standard algorithm for minimizing regret is the 
\emph{multiplicative weights update method}. We recover this by letting $f(x,z) = x + \log(1-z)$ 
and $\psi(w) = \sum_{i=1}^n w_i\log w_i$. Then $\psi^*(x) = \log\left(\sum_{i=1}^n \exp(x_i)\right)$, 
and $w_{t+1,i} = \exp(x_{t+1,i})/\sum_{j=1}^n \exp(x_{t+1,j})$.
\begin{align*}
\psi^*(f(x,\eta z)) &= \log\left(\sum_{i=1}^n \exp(x_i + \log(1-\eta z_i))\right) \\
 &= \log\left(\sum_{i=1}^n \exp(x_i)(1-\eta z_i)\right) \\
 &= \log\left(\sum_{i=1}^n \exp(x_i) - \eta \sum_{i=1}^n \exp(x_i) z_i\right) \\
 &\leq \log\left(\sum_{i=1}^n \exp(x_i)\right) - \eta \frac{\sum_{i=1}^n \exp(x_i) z_i}{\sum_{i=1}^n \exp(x_i)} \\
 &= \psi^*(x) - \eta w^Tz,
\end{align*}
so condition 1 is satisfied.

@@

\end{document}

>>>>

\title{Online Learning Framework}
\author{Jacob Steinhardt}
\begin{document}
\maketitle
We are interested in the following learning problem: our experts 
consist of vectors $w \in \sH$, where $\sH$ is some convex 
set; similarly, our adversary can play any move $z_t$ with 
$z_t \in \sA$. A typical example is $\sH$ is the $n$-dimensional 
simplex and $\sA$ is the unit $l^{\infty}$ ball. Our end goal is 
to obtain a regret bound of the following type:
\[ \sum_{t=1}^T w_t^Tz_t \leq \frac{\rho}{\epsilon} + \epsilon \sum_{t=1}^T |w|^T|z_t| + \sum_{t=1}^T w^Tz_t. \]
If we let $\sigma = \sup_{w \in \sH, z \in \sA} |w|^T|z|$, then the regret is bounded by
$\frac{\rho}{\epsilon} + \epsilon \sigma T$, which is equal to $2\sqrt{\rho \sigma T}$ for 
$\epsilon = \sqrt{\frac{\rho}{\sigma T}}$. This is the type of regret bound that is typical 
in most online learning settings. However, our bound can lead to stronger results in some 
situations, for instance if $\sA$ is very asymmetrical about the origin, in which case 
$|w|^T|z_t|$ will tend to be close to $w^Tz_t$ and we get bound
\[ \sum_{t=1}^T w_t^Tz_t \leq \frac{\rho}{\epsilon} + (1+\epsilon)\sum_{t=1}^T w^Tz_t, \]
which is potentially much better, for instance if there is an expert that achieves 
very small loss.

We propose the following general style of algorithm for obtaining such 
bounds. The key innovation is an auxiliary variable $x_t$ that lets us 
relate $w_t^Tz_t$ and $w^Tz_t$. The updates are as follows:
\begin{align*}
x_{t+1} &= f(x_t, \eta z_t) \\
w_{t+1} &= \argmax_{w \in \sH} w^Tx_{t+1} - \psi(w) \\
\Phi_{t+1} &= w_{t+1}^Tx_{t+1} - \psi(w_{t+1})  \\ &= \psi^*(x_{t+1}).
\end{align*}

@@

\paragraph{Formal statement.} To turn these approximations into a rigorous bound, we will 
need $f$ and $\psi$ to satisfy the following properties:
\begin{itemize}
\item Concavity of $\psi^* \circ f$: $\psi^*(f(x,\eta z)) \leq \psi^*(x) - \eta w_x^Tz$, where $w_x = \partial \psi^*(x)$.
\item Smoothness of $f$: $w^T(x-\eta z - f(x,\eta z)) \leq \eta^2 \alpha |w|^T|z|$ for all $w$.
\end{itemize}
(For some intuition, any $f$ satisfying the two properties above 
must satisfy $f(x,0) = x$ and $f(x,\eta z) = x-\eta z + O(\eta^2)$.)
\begin{theorem}
\label{thm:main}
Suppose that $\psi^*$ and $f$ satisfy the two properties above. Then, 
for $\eta = \frac{\epsilon}{\alpha}$, we have the bound
\[ \sum_{t=1}^T w_t^Tz_t \leq \alpha \frac{\Phi_1 + \psi(w) - w^Tx_1}{\epsilon} + \epsilon \sum_{t=1}^T |w|^T|z_t| + \sum_{t=1}^T w^Tz_t. \]
\end{theorem}
\begin{proof}
By the first condition, we have
\begin{align*}
\Phi^{t+1} &= \psi^*(x_{t+1}) \\
 &= \psi^*(f(x_t, \eta z_t)) \\
 &\leq \psi^*(x_t) - \eta \left(\frac{\partial \psi^*}{\partial x_t}\right)^T z_t \\
 &= \Phi_t - \eta w_t^T z_t,
\end{align*}
which by induction implies that
\[ \Phi_{T+1} \leq \Phi_1 - \eta \sum_{t=1}^T w_t^Tz_t. \]
By the second condition, we have
\begin{align*}
w^Tx_{t+1} &= w^Tf(x, \eta z) \\
 &\geq w^Tx_t - \eta w^Tz_t - \eta^2 \alpha |w|^T|z_t|,
\end{align*}
which by induction implies that
\[ w^Tx_{T+1} \geq w^Tx_1 - \eta^2\alpha \sum_{t=1}^T |w|^T|z_t| - \eta \sum_{t=1}^T w^Tz_t. \]
Combining this with the first inequality yields
\begin{align*}
\Phi_1 - \eta \sum_{t=1}^T w_t^Tz_t &\geq \Phi_{T+1} \\
 &\geq w^Tx_{T+1} - \psi(w) \\
 &\geq w^Tx_1 - \psi(w) - \eta^2\alpha \sum_{t=1}^T |w|^T|z_t| - \eta \sum_{t=1}^T w^Tz_t.
\end{align*}
Re-arranging yields
\[ \sum_{t=1}^T w_t^Tz_t \leq \frac{\Phi_1 + \psi(w) - w^Tx_1}{\eta} + \eta\alpha \sum_{t=1}^T |w|^T|z_t| + \sum_{t=1}^T w^Tz_t, \]
which gives the desired result.
\end{proof}

\paragraph{Relationship to mirror descent.} Mirror descent can be cast in the same framework as above, 
and corresponds to the exact additive update $x_{t+1} = x_t - \eta z_t$. Thus while our algorithm 
makes sure that $\psi^*$ decreases at a rate of $w_t^Tz_t$ and then bounds the error between 
$x_{t+1}$ and $x_t - \eta z_t$, in contrast mirror descent sets the error between $x_{t+1}$ and 
$x_t - \eta z_t$ to zero, and then relies on the smoothness of $\psi^*$ to bound the error between 
$\psi^*(x-\eta z)$ and $\psi^*(x) - \eta w_x^Tz$.
\section{Examples}
\paragraph{Multiplicative updates.} If we let $\sH$ be the simplex and $\sA$ be the 
$l^{\infty}$ ball of radius $r$, then a standard algorithm for minimizing regret is the 
\emph{multiplicative weights update method}. We recover this by letting $f(x,z) = x + \log(1-z)$ 
and $\psi(w) = \sum_{i=1}^n w_i\log w_i$. Then $\psi^*(x) = \log\left(\sum_{i=1}^n \exp(x_i)\right)$, 
and $w_{t+1,i} = \exp(x_{t+1,i})/\sum_{j=1}^n \exp(x_{t+1,j})$.
\begin{align*}
\psi^*(f(x,\eta z)) &= \log\left(\sum_{i=1}^n \exp(x_i + \log(1-\eta z_i))\right) \\
 &= \log\left(\sum_{i=1}^n \exp(x_i)(1-\eta z_i)\right) \\
 &= \log\left(\sum_{i=1}^n \exp(x_i) - \eta \sum_{i=1}^n \exp(x_i) z_i\right) \\
 &\leq \log\left(\sum_{i=1}^n \exp(x_i)\right) - \eta \frac{\sum_{i=1}^n \exp(x_i) z_i}{\sum_{i=1}^n \exp(x_i)} \\
 &= \psi^*(x) - \eta w^Tz,
\end{align*}
so condition 1 is satisfied.

@@

\paragraph{Sum of pointed convex sets.} Suppose that our hypothesis class 
$\sH \subseteq \bR^n$ can be decomposed as $\sH = \sH_+ + \sH_-$, where $\sH_+$ consists 
only of non-negative vectors and $\sH_-$ consists only of non-positive vectors. 
We can translate this into a problem n $\bR^{2n}$, by letting $\sH' \eqdef \sH_+ \times (-\sH_-)$ 
and $\sA' \eqdef \{(z, -z) \mid z \in \sA\}$. Then $(w')^T(z') = w^Tz$. Now 
let $(w_+, w_-)$ be the decomposition of $w'$ into its two components. We will 
define
\[ \psi(w') = \left\{ \begin{array}{ccl} \sum_{i=1}^n \log((w_+)_i + (w_-)_i) & : & w' \in \sH \\ \infty & : & \mathrm{else}. \end{array} \right. \]
Now let 
\[ R \eqdef \sup\limits_{w_+ \in \sH_+, w_- \in \sH_-, z \in \sA}  |w_+^Tz| + |w_-^Tz|, \]
and consider the update $x_{t+1} = x_t - \eta z'_t - 2\eta^2 R |z'_t|$. Then we have
\begin{align*}
\psi^*(x_{t+1}) &= \psi^*(x_t) - \eta (w'_t)^T(z'_t) - 2\eta^2R (w'_t)^T|z'_t| + D_{\psi^*}(x_{t+1} \| x_t) \\
 &= \psi^*(x_t) - \eta (w'_t)^Tz_t - 2\eta^2R (w'_t)^T|z'_t| + \frac{1}{2}((w'_t)^T|x_{t+1}-x_t|)^2 \\
 &= \psi^*(x_t) - \eta (w'_t)^Tz_t - 2\eta^2R (w'_t)^T|z'_t| + \frac{1}{2}(\eta (w'_t)^T|z'_t| + 2\eta^2R (w'_t)^T|z'_t|)^2 \\
 &= \psi^*(x_t) - \eta (w'_t)^Tz_t
\end{align*}
as long as $2\eta R \leq 1$, or $\eta \leq \frac{1}{2R}$. Thus the first condition is 
satisfied. Also, by construction we have 
\[ w^T(x_t - \eta z_t - x_{t+1}) = 2\eta^2 R w^T|z'_t|, \]
so that the second condition is satisfied for $\alpha = 2R$. Combining these, and 
setting $x_1 = 0$, we obtain the bound
\[ \sum_{t=1}^T w_t^Tz_t \leq \frac{2R[\psi(w) - \inf_{\tilde{w}} \psi(\tilde{w})]}{\epsilon} + \epsilon \sum_{t=1}^T (w_+ + w_-)^T|z_t| + \sum_{t=1}^T w^Tz_t. \]
This is a quite general result, but suffers from the fact that $\psi(w)$ can grow very large. 
Finding ways to deal with this would thus be very interesting.

\paragraph{Inflated $l^p$ norm.} By abuse of notation, let $\alpha$ denote the vector 
whose coordinates are all equal to $\alpha$. Let $\sH = B_{\alpha}^p(\rho)$, where 
\[ B_{\alpha}^p(\rho) \eqdef \{w \mid w \geq \alpha, \|w-\alpha\|_p \leq \rho \}. \]
Note that this is the positive $l^p$ ball of radius $\rho$, translated by $\alpha$. 
(We will show below that we can embed the unconstrained $l^p$ ball into $B_{\alpha}^p(\rho)$ 
at the cost of increasing the dimension by a factor of 2, implying that 
$B_{\alpha}^p(\rho)$ is a rich hypothesis class.)

We will let $\psi(w) = \left\{ \begin{array}{ccl} \frac{1}{2(p-1)}\|w-\alpha\|_p^2 & : & w \geq \alpha \\ \infty & : & w \not\geq \alpha. \end{array} \right.$. 
Then 
\[ \psi^*(x) = \alpha \sum_{i=1}^n x_i + \frac{1}{2(q-1)}\|x_+\|_q^2, \]
where $x_+$ is the positive part of $x$ and $\frac{1}{p} + \frac{1}{q} = 1$.

Now suppose we use the update $x_{t+1,i} = x_{t,i} - \eta z_{t,i} - 2\eta^2 \frac{\|z\|_q^2}{\|z\|_1} |z_{t,i}| / \alpha$. 
Note that the Bregman divergence $D_{\psi^*}(x_{t+1} \| x_t)$ is bounded above by the Bregman divergence 
of $\frac{1}{2(q-1)}\|x\|_q^2$ (since it can be obtained from $\psi^*$ by adding a convex function), 
which is in turn bounded above by $\frac{1}{2}\|x_{t+1}-x_t\|_q^2$. Using these facts, together with the fact that 
$w \geq \alpha$, we have
\begin{align*}
\psi^*(x_{t+1}) &\leq \psi^*(x_t) - \eta w_t^Tz_t - \frac{2\eta^2\|z\|_q^2}{\alpha \|z\|_1} w_t^T|z_t| + D_{\psi^*}(x_{t+1} \| x_t) \\
 &\leq \psi^*(x_t) - \eta w_t^Tz_t - \frac{2\eta^2\|z\|_q^2}{\alpha \|z\|_1} w_t^T|z_t| + \frac{1}{2}\left\|\eta z_t + \frac{2\eta^2\|z\|_q^2}{\alpha \|z\|_1} |z_t|\right\|_q^2 \\
 &\leq \psi^*(x_t) - \eta w_t^Tz_t - 2\eta^2 \|z\|_q^2 + \frac{1}{2} \left(\eta \|z_t\|_q + \frac{2\eta^2\|z_t\|_q^3}{\alpha \|z_t\|_1}\right)^2 \\
 &\leq \psi^*(x_t) - \eta w_t^Tz_t - 2\eta^2 \|z\|_q^2 + \frac{1}{2} \left(\eta \|z_t\|_q + 2\eta^2 \|z_t\|_{q}^2 / \alpha\right)^2 \\
 &\leq \psi^*(x_t) - \eta w_t^Tz_t
\end{align*}
as long as $2\eta \|z_t\|_q/\alpha \leq 1$, which holds as long as $\eta \leq \frac{\alpha}{2r_q}$. In addition, we have that
\begin{align*}
w^T(x_{t} - \eta z_t - x_{t+1}) &= -\frac{2\eta^2\|z_t\|_q^2}{\alpha\|z_t\|_1}w^T|z_t| \\
 &\leq \eta^2 \frac{2\|z_t\|_q}{\alpha} |w|^T|z_t| \\
 &\leq \eta^2 \frac{2r_q}{\alpha} |w|^T|z_t|.
\end{align*}
So, both conditions of Theorem~\ref{thm:main} are satisfied. 
If we set $x_1 = 0$, then $\psi^*(x_1) = 0$ as well.
Therefore, for $\eta = \frac{\epsilon \alpha}{2r_q}$, we obtain a regret bound of 
\[ \frac{2r_q\psi(w)}{\epsilon \alpha} + \epsilon \sum_{t=1}^T |w|^T|z_t|. \]
Note that the regret gets smaller as $\alpha$ grows larger. This might seem too good 
to be true, until we notice the following: if $\alpha$ is very large, then all 
vectors with a small value of $\|w-\alpha\|_p$ will point in approximately the same 
direction, and thus get similar (multiplicative) values for the regret.
\paragraph{Embedding the $l^p$ ball into $B_{\alpha}^p$.} We now show how to turn problems 
involving the $l^p$ ball into problems involving $B_{\alpha}^p$. Given a vector 
$w \in \bR^n$, we can create a vector $w' \in \bR^{2n}$ via the embedding 
$w \mapsto \left[ \begin{array}{c} \alpha + \max(w, 0) \\ \alpha - \min(w,0) \end{array} \right]$. 
This embedding satisfies the property that $\|w\|_p = \|w'-\alpha\|_p$. Also, $w^Tz = w'^T\left[ \begin{array}{c} z \\ -z \end{array} \right]$. 
If we plug into the regret bound above, the only difference is that the value of $|w|^T|z_t|$ increases, yielding 
a regret of
\[ \frac{r_q \|w\|_p^2}{\epsilon \alpha (p-1)} + \epsilon \sum_{t=1}^T \left(|w|^T|z_t| + 2\alpha \|z_t\|_1\right). \]
Unfortunately, this final bound is quite bad, because the $\|z_t\|_1$ term grows with $T$ and depends on the 
$l^1$ norm, which is the biggest of all the $l^p$ norms.
\end{document}
